---
title: "Algorithm comparison"
author: "Cristian Castiglione"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ants}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Workspace setup

```{r setup, include = FALSE}
options(warn = -1)
options(rmarkdown.html_vignette.check_title = FALSE)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Load the \code{sgdGMF} package in the workspace.

```{r sgdgmf, results = FALSE, message = FALSE, warning = FALSE}
library(sgdGMF)
# devtools::load_all()
```

Load other useful packages in the workspace.

```{r libraries, results = FALSE, message = FALSE, warning = FALSE}
library(ggplot2)
library(ggpubr)
library(reshape2)
library(viridis)

theme_set(theme_bw())
```

## Ant traits data

Load the ant traits data in the workspace and define the response matrix `Y` and covariate matrices `X` and `Z`.

```{r data}
data(antTraits, package = "mvabund")

Y = as.matrix(antTraits$abund)
X = as.matrix(antTraits$env[,-3])
Z = matrix(1, nrow = ncol(Y), ncol = 1)

n = nrow(Y)
m = ncol(Y)
```

## Model specification

Set the model family to Poisson since the response matrix contain count data.

```{r family}
family = poisson()
```

Select the optimal number of latent factors using the function \code{sgdgmf.rank},
which employs an adjusted eigenvalue thresholding method to identify the optimal
elbow point of a screeplot.

```{r rank}
ncomp = sgdgmf.rank(Y = Y, X = X, Z = Z, family = family)$ncomp
cat("Selected rank: ", ncomp)
```

## Model estimation

Estimate a Poisson GMF model using iterated least squares, diagonal quasi-Newton and
stochastic gradient descent algorithms.

```{r fit}
gmf.airwls = sgdgmf.fit(Y, X, Z, ncomp = ncomp, family = family, method = "airwls")
gmf.newton = sgdgmf.fit(Y, X, Z, ncomp = ncomp, family = family, method = "newton")
gmf.sgd    = sgdgmf.fit(Y, X, Z, ncomp = ncomp, family = family, method = "sgd", sampling = "block")
```

## Model validation

Compute the correlation between the observed data and the predicted values.

```{r cor}
cat(" AIRWLS: ", 100 * round(cor(c(Y), c(gmf.airwls$mu)), 4), "\n",
    " Newton: ", 100 * round(cor(c(Y), c(gmf.newton$mu)), 4), "\n",
    " SGD:    ", 100 * round(cor(c(Y), c(gmf.sgd$mu)), 4), sep = "")
```
Compute the partial deviance residuals holding out from the model the estimated 
matrix factorization. Additionally, compute the spectrum of such a residual matrix.

```{r resid}
res.airwls = residuals(gmf.airwls, partial = FALSE, spectrum = TRUE, ncomp = 20)
res.newton = residuals(gmf.newton, partial = FALSE, spectrum = TRUE, ncomp = 20)
res.sgd = residuals(gmf.sgd, partial = FALSE, spectrum = TRUE, ncomp = 20)
```

Analyze the residual distribution.

```{r hist, fig.width = 6, fig.height = 5}
get.factor = function (levels, nrep) factor(rep(levels, each = nrep), levels = levels)

df = data.frame(
  residuals = c(res.airwls$residuals, res.newton$residuals, res.sgd$residuals),
  fitted = c(gmf.airwls$mu, gmf.newton$mu, gmf.sgd$mu),
  model = get.factor(c("AIRWLS", "Newton", "SGD"), n)
)

plt.res.vs.fit = ggplot(data = df, map = aes(x = fitted, y = residuals)) +
  geom_point(alpha = 0.5) + facet_grid(cols = vars(model)) + 
  geom_hline(yintercept = 0, col = 2, lty = 2)

plt.res.hist = ggplot(data = df, map = aes(x = residuals, y = after_stat(density))) +
  geom_histogram(bins = 30) + facet_grid(cols = vars(model)) + 
  geom_vline(xintercept = 0, col = 2, lty = 2)

ggpubr::ggarrange(plt.res.vs.fit, plt.res.hist, nrow = 2, align = "v")

```

Plot the variance explained by each principal component of the residual matrix.

```{r spectrum, fig.width = 6, fig.height = 2.5}
data.frame(
  components = rep(1:20, 3),
  variance = c(res.airwls$lambdas, res.newton$lambdas, res.sgd$lambdas),
  model = get.factor(c("AIRWLS", "Newton", "SGD"), 20)) |>
  ggplot(map = aes(x = components, y = variance)) +
  geom_col(color = "white") + facet_grid(cols = vars(model))
```

## Observations vs fitted values

Plot the abundance predicted by each method comparing it with the observed matrix.

```{r pred, fig.width = 6, fig.height = 5}
colnames(gmf.airwls$mu) = colnames(Y)
colnames(gmf.newton$mu) = colnames(Y)
colnames(gmf.sgd$mu) = colnames(Y)

df = rbind(
  reshape2::melt(Y, varnames = c("environments", "species"), value.name = "abundance"),
  reshape2::melt(gmf.airwls$mu, varnames = c("environments", "species"), value.name = "abundance"),
  reshape2::melt(gmf.newton$mu, varnames = c("environments", "species"), value.name = "abundance"),
  reshape2::melt(gmf.sgd$mu, varnames = c("environments", "species"), value.name = "abundance"))

df$model = get.factor(c("Data", "AIRWLS", "Newton", "SGD"), n*m)

ggplot(data = df, map = aes(x = species, y = environments, fill = abundance)) + 
  geom_raster() + facet_wrap(vars(model), nrow = 2, ncol = 2) + scale_fill_viridis() + 
  theme(axis.text = element_blank(), axis.ticks = element_blank(), panel.grid = element_blank()) +
  labs(x = "Species", y = "Environmens", fill = "Abundance", title = "Ant abundance data")
```

## Latent scores and low-dimensional representation

```{r scores, fig.width = 8, fig.height = 5}
pca.airwls = RSpectra::svds(tcrossprod(gmf.airwls$U, gmf.airwls$V), 2)
pca.newton = RSpectra::svds(tcrossprod(gmf.newton$U, gmf.newton$V), 2)
pca.sgd = RSpectra::svds(tcrossprod(gmf.sgd$U, gmf.sgd$V), 2)

plt.envs = data.frame(
  idx = rep(1:n, times = 3),
  pc1 = c(scale(pca.airwls$u[,1]), scale(pca.newton$u[,1]), scale(pca.sgd$u[,1])),
  pc2 = c(scale(pca.airwls$u[,2]), scale(pca.newton$u[,2]), scale(pca.sgd$u[,2])),
  model = get.factor(c("AIRWLS", "Newton", "SGD"), n)) |>
  ggplot(map = aes(x = pc1, y = pc2, color = idx, label = idx)) +
  geom_hline(yintercept = 0, lty = 2, color = "grey40") + 
  geom_vline(xintercept = 0, lty = 2, color = "grey40") + 
  geom_point() + facet_grid(cols = vars(model)) + 
  scale_color_viridis(option = "viridis") +
  geom_text(color = 1, size = 2.5, nudge_x = -0.2, nudge_y = +0.2) +
  labs(x = "PC 1", y = "PC 2", color = "Environments")

plt.species = data.frame(
  idx = rep(1:m, times = 3),
  pc1 = c(scale(pca.airwls$v[,1]), scale(pca.newton$v[,1]), scale(pca.sgd$v[,1])),
  pc2 = c(scale(pca.airwls$v[,2]), scale(pca.newton$v[,2]), scale(pca.sgd$v[,2])),
  model = get.factor(c("AIRWLS", "Newton", "SGD"), m)) |>
  ggplot(map = aes(x = pc1, y = pc2, color = idx, label = idx)) +
  geom_hline(yintercept = 0, lty = 2, color = "grey40") + 
  geom_vline(xintercept = 0, lty = 2, color = "grey40") + 
  geom_point() + facet_grid(cols = vars(model)) + 
  scale_color_viridis(option = "inferno") +
  geom_text(color = 1, size = 2.5, nudge_x = -0.2, nudge_y = +0.2) +
  labs(x = "PC 1", y = "PC 2", color = "Species")

ggpubr::ggarrange(plt.envs, plt.species, nrow = 2, ncol = 1, align = "hv")
```


