---
title: "Ant traits data"
author: "Cristian Castiglione"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ants}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Workspace setup

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Load the \code{sgdGMF} package in the workspace.

```{r sgdgmf}
# library(sgdGMF)
devtools::load_all()
```

Load other useful packages in the workspace.

```{r libraries}
library(mvabund)
library(ggplot2)
library(ggpubr)
library(reshape2)
library(viridis)
```

## Ant traits data

Load the ant traits data in the workspace and define the response and covariate matrices.

```{r data}
data(antTraits)

Y = as.matrix(antTraits$abund)
X = as.matrix(antTraits$env[,-3])
Z = matrix(1, nrow = ncol(Y), ncol = 1)

n = nrow(Y)
m = ncol(Y)
```

## Model specification

Set the model family to Poisson since the response matrix contain count data.

```{r family}
family = poisson()
```

Select the optimal number of latent factors using the function \code{sgdgmf.rank},
which employs an adjusted eigenvalue thresholding method to identify the optimal
elbow point of a screeplot.

```{r rank}
ncomp = sgdgmf.rank(Y = Y, X = X, Z = Z, family = family)$ncomp
cat("Selected rank: ", ncomp)
```

## Model estimation

Estimate a Poisson GMF model using iterated least squares, diagonal quasi-Newton and
stochastic gradient descent algorithms.

```{r fit}
ctr = list(verbose = FALSE)

gmf.airwls = sgdgmf.fit(Y, X, Z, ncomp = ncomp, family = family, method = "airwls", control.alg = ctr)
gmf.newton = sgdgmf.fit(Y, X, Z, ncomp = ncomp, family = family, method = "newton", control.alg = ctr)
gmf.bsgd = sgdgmf.fit(Y, X, Z, ncomp = ncomp, family = family, method = "bsgd", control.alg = ctr)
```

## Model validation

Compute the correlation between the observed data and the predicted values.

```{r cor}
{
cat("\n AIRWLS: ", 100 * round(cor(c(Y), c(gmf.airwls$mu)), 4))
cat("\n Newton: ", 100 * round(cor(c(Y), c(gmf.newton$mu)), 4))
cat("\n B-SGD:  ", 100 * round(cor(c(Y), c(gmf.bsgd$mu)), 4))
}
```
Compute the partial deviance residuals holding out from the model the estimated 
matrix factorization. Additionally, compute the spectrum of such a residual matrix.

```{r resid}
eig.airwls = residuals(gmf.airwls, partial = FALSE, spectrum = TRUE, ncomp = 20)
eig.newton = residuals(gmf.newton, partial = FALSE, spectrum = TRUE, ncomp = 20)
eig.bsgd = residuals(gmf.bsgd, partial = FALSE, spectrum = TRUE, ncomp = 20)
```

Analyze the residual distribution.

```{r hist, fig.width = 6, fig.height = 5}
get.factor = function (levels, nrep) factor(rep(levels, each = nrep), levels = levels)

df = data.frame(
  residuals = c(eig.airwls$residuals, eig.newton$residuals, eig.bsgd$residuals),
  fitted = c(gmf.airwls$mu, gmf.newton$mu, gmf.bsgd$mu),
  model = get.factor(c("AIRWLS", "Newton", "SGD"), n)
)

plt.res.vs.fit = ggplot(data = df, map = aes(x = fitted, y = residuals)) +
  geom_point(alpha = 0.5) + facet_grid(cols = vars(model)) + 
  geom_hline(yintercept = 0, col = 2, lty = 2) + theme_grey()

plt.res.hist = ggplot(data = df, map = aes(x = residuals, y = after_stat(density))) +
  geom_histogram(bins = 30) + facet_grid(cols = vars(model)) + 
  geom_vline(xintercept = 0, col = 2, lty = 2) + theme_grey()

ggpubr::ggarrange(plt.res.vs.fit, plt.res.hist, nrow = 2)

```

Plot the variance explained by each principal component of the residual matrix.

```{r spectrum, fig.width = 6, fig.height = 2.5}
df = data.frame(
  components = rep(1:20, 3),
  variance = c(eig.airwls$spectrum / eig.airwls$total.var, 
               eig.newton$spectrum / eig.newton$total.var, 
               eig.bsgd$spectrum / eig.bsgd$total.var),
  model = get.factor(c("AIRWLS", "Newton", "SGD"), 20)
)

ggplot(data = df, map = aes(x = components, y = variance)) +
  geom_col() + facet_grid(cols = vars(model)) + theme_grey()
```

## Observations vs fitted values

Plot the abundance predicted by each method comparing it with the observed matrix.

```{r pred, fig.width = 6, fig.height = 5}
colnames(gmf.airwls$mu) = colnames(Y)
colnames(gmf.newton$mu) = colnames(Y)
colnames(gmf.bsgd$mu) = colnames(Y)

df = rbind(
  reshape2::melt(Y, varnames = c("environments", "species"), value.name = "abundance"),
  reshape2::melt(gmf.airwls$mu, varnames = c("environments", "species"), value.name = "abundance"),
  reshape2::melt(gmf.newton$mu, varnames = c("environments", "species"), value.name = "abundance"),
  reshape2::melt(gmf.bsgd$mu, varnames = c("environments", "species"), value.name = "abundance"))

df$model = get.factor(c("Data", "AIRWLS", "Newton", "SGD"), n*m)

ggplot(data = df, map = aes(x = species, y = environments, fill = abundance)) + 
  geom_raster() + facet_wrap(vars(model), nrow = 2, ncol = 2) + 
  scale_fill_viridis() + theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  labs(x = "Species", y = "Environmens", fill = "Abundance", title = "Ant abundance")
```

## Latent scores and low-dimensional representation

```{r scores, fig.width = 6, fig.height = 2.5}
df = data.frame(
  idx = rep(1:n, times = 3),
  pc1 = c(gmf.airwls$U[,1], gmf.newton$U[,1], gmf.bsgd$U[,1]),
  pc2 = c(gmf.airwls$U[,2], gmf.newton$U[,2], gmf.bsgd$U[,2]),
  model = get.factor(c("AIRWLS", "Newton", "SGD"), n)
)

ggplot(data = df, map = aes(x = pc1, y = pc2, color = idx)) +
  geom_point() + facet_grid(cols = vars(model)) + 
  scale_color_viridis() + theme_grey() +
  labs(x = "PC 1", y = "PC 2", color = "Index")
```





