---
title: "Introduction to the sgdGMF package"
author: "Cristian Castiglione"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{intro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Workspace setup

```{r setup, include = FALSE}
options(warn = -1)
options(rmarkdown.html_vignette.check_title = FALSE)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Load the \code{sgdGMF} package in the workspace.

```{r sgdgmf}
# library(sgdGMF)
devtools::load_all()
```

Load other useful packages in the workspace.

```{r libraries}
library(ggplot2)
library(ggpubr)
library(reshape2)
library(viridis)

theme_set(theme_bw())
```

## Ant traits data

Load the ant traits data in the workspace and define the response and covariate matrices.

```{r data}
data(antTraits, package = "mvabund")

Y = as.matrix(antTraits$abund)
X = as.matrix(antTraits$env[,-3])
Z = matrix(1, nrow = ncol(Y), ncol = 1)

n = nrow(Y)
m = ncol(Y)
```

## Model specification

Set the model family to Poisson since the response matrix contain count data.

```{r family}
family = poisson()
```

Select the optimal number of latent factors using the function \code{sgdgmf.rank},
which employs an adjusted eigenvalue thresholding method to identify the optimal
elbow point of a screeplot.

```{r rank}
vglm = sgdgmf.rank(Y = Y, X = X, Z = Z, family = family, method = "onatski",
                   return.eta = TRUE, return.mu = TRUE, return.res = TRUE)

cat("Selected rank: ", vglm$ncomp)
```

## Model estimation

Estimate a Poisson GMF model using iterated least squares.

```{r fit}
ncomp = vglm$ncomp
gmf = sgdgmf.fit(Y, X, Z, ncomp = ncomp, family = family, method = "bsgd")
```

## Model validation

Compute the correlation between the observed data and the predicted values.

```{r cor}
cat(" VGLM: ", 100 * round(cor(c(Y), c(vglm$mu)), 4), "\n",
    "  GMF: ", 100 * round(cor(c(Y), c(gmf$mu)), 4), sep = "")
```
Compute the partial deviance residuals holding out from the model the estimated 
matrix factorization. Additionally, compute the spectrum of such a residual matrix.

```{r resid}
res = residuals(gmf, spectrum = TRUE, ncomp = 20)
```

Compare the residuals of two competing models: VGLM and GMF.
Notice that VGLM is a particular case of GMF of which only include the regression
effects and does not include a residual matrix factorization in the linear predictor.

```{r hist, fig.width = 7, fig.height = 5}
plt.res.fit.glm = plot(gmf, type = "fit", partial = TRUE)
plt.res.hist.glm = plot(gmf, type = "hist", partial = TRUE)
plt.res.fit.gmf = plot(gmf, type = "fit", partial = FALSE)
plt.res.hist.gmf = plot(gmf, type = "hist", partial = FALSE)

ggpubr::ggarrange(
  plt.res.fit.glm + ggtitle("Residuals vs Fitted values (VGLM)"), 
  plt.res.hist.glm + ggtitle("Histogram of the residuals (VGLM)"), 
  plt.res.fit.gmf + ggtitle("Residuals vs Fitted values (GMF)"), 
  plt.res.hist.gmf + ggtitle("Histogram of the residuals (GMF)"), 
  nrow = 2, ncol = 2, align = "hv")

```

Plot the variance explained by each principal component of the residual matrix.

```{r spectrum, fig.width = 7, fig.height = 3}
plt.eig.glm = screeplot(gmf, partial = TRUE) + ggtitle("Residual screeplot (VGLM)")
plt.eig.gmf = screeplot(gmf, partial = FALSE) + ggtitle("Residual screeplot (GMF)")

ggpubr::ggarrange(plt.eig.glm, plt.eig.gmf, nrow = 1, ncol = 2, align = "hv")
```

## Observations vs fitted values

Plot the abundance predicted by each method comparing it with the observed matrix.

```{r pred, fig.width = 7, fig.height = 3.5}
plt.ant = image(gmf, limits = range(c(Y)), type = "data")
plt.fit = image(gmf, limits = range(c(Y)), type = "response")

ggpubr::ggarrange(
  plt.ant + labs(x = "Species", y = "Environments", title = "Observed abundance"), 
  plt.fit + labs(x = "Species", y = "Environments", title = "Predicted abundance"), 
  nrow = 1, ncol = 2, common.legend = TRUE, legend = "bottom", align = "hv")
```

```{r resid2, fig.width = 7, fig.height = 3.5}
plt.dev = image(gmf, type = "deviance", resid = TRUE, symmetric = TRUE)
plt.prs = image(gmf, type = "pearson", resid = TRUE, symmetric = TRUE)

ggpubr::ggarrange(
  plt.dev + labs(x = "Species", y = "Environments", title = "Deviance residuals"), 
  plt.prs + labs(x = "Species", y = "Environments", title = "Pearson residuals"), 
  nrow = 1, ncol = 2, common.legend = FALSE, legend = "bottom", align = "hv")
```

## Latent scores and low-dimensional representation

```{r scores, fig.width = 7, fig.height = 4}
plt = biplot(gmf)
ggpubr::ggarrange(
  plt$scores + ggtitle("Environments"), 
  plt$loadings + ggtitle("Species"), 
  nrow = 1, ncol = 2, align = "hv")
```


