---
title: "Analysis of the residuals"
author: "Cristian Castiglione"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{intro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Workspace setup

```{r setup, include = FALSE}
options(warn = -1)
options(rmarkdown.html_vignette.check_title = FALSE)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Load the \code{sgdGMF} package in the workspace.

```{r sgdgmf}
library(sgdGMF)
```

Load other useful packages in the workspace.

```{r libraries}
library(ggplot2)
library(ggpubr)
library(reshape2)

theme_set(theme_bw())
```

Install the \code{mvabund} package.

```{r datapkg}
if (!require("mvabund", quietly = TRUE)) {
  install.packages("RcppGSL", repos = "https://cloud.r-project.org")
  install.packages("mvabund", repos = "https://cloud.r-project.org") 
}
```

## Ant traits data

Load the ant traits data in the workspace and define the response matrix `Y` and covariate matrices `X` and `Z`.

```{r data}
data(antTraits, package = "mvabund")

Y = as.matrix(antTraits$abund)
X = as.matrix(antTraits$env[,-3])
Z = matrix(1, nrow = ncol(Y), ncol = 1)

n = nrow(Y)
m = ncol(Y)
```

## Model specification

Set the model family to Poisson since the response matrix contain count data.

```{r family}
family = poisson()
```

Select the optimal number of latent factors using the function \code{sgdgmf.rank},
which employs an adjusted eigenvalue thresholding method to identify the optimal
elbow point of a screeplot.

```{r rank}
ncomp = sgdgmf.rank(Y = Y, X = X, Z = Z, family = family)$ncomp
cat("Selected rank: ", ncomp)
```

## Model estimation

Estimate a Poisson GMF model using iterated least squares.

```{r fit}
gmf = sgdgmf.fit(Y, X, Z, ncomp = ncomp, family = family, method = "airwls")
```

## Model validation

Compute the deviance residuals of the model the estimated matrix factorization. 
Additionally, compute the spectrum of such a residual matrix.

```{r resid}
res = residuals(gmf, spectrum = TRUE, ncomp = 20)
```

Compare the residuals of two competing models: VGLM and GMF.
Notice that VGLM is a particular case of GMF of which only include the regression
effects and does not include a residual matrix factorization in the linear predictor.

```{r plot, fig.width = 7, fig.height = 5}
ggpubr::ggarrange(
  plot(gmf, type = "res-idx"),
  plot(gmf, type = "res-fit"),
  plot(gmf, type = "hist"),
  plot(gmf, type = "qq"),
  nrow = 2, ncol = 2, align = "hv")
```

We now have a look to the spectrum of the residual matrices, i.e., the eigenvalues
of the corresponding covariance matrix. However, instead of analyzing the actual
values of the eigenvalues, we normalize them in such a way to plot the percentage of
variance explained by each principal component.

```{r spectrum, fig.width = 7, fig.height = 3}
ggpubr::ggarrange(
  screeplot(gmf, cumulative = FALSE, proportion = TRUE),
  screeplot(gmf, cumulative = TRUE, proportion = TRUE),
  nrow = 1, ncol = 2, align = "hv")
```

## Observations vs fitted values

Plot the deviance and Pearson residuals using a heatmap. This could be helpful to
graphically detect if there are some structured patterns in the matrix that have not 
been captured by the model.

```{r resid2, fig.width = 7, fig.height = 3.5}
plt.dev = image(gmf, type = "deviance", resid = TRUE, symmetric = TRUE)
plt.prs = image(gmf, type = "pearson", resid = TRUE, symmetric = TRUE)

ggpubr::ggarrange(
  plt.dev + labs(x = "Species", y = "Environments", title = "Deviance residuals"), 
  plt.prs + labs(x = "Species", y = "Environments", title = "Pearson residuals"),
  nrow = 1, ncol = 2, common.legend = FALSE, legend = "bottom", align = "hv")
```



